---
title: "HW 10"
author: "Davis Vaughan"
date: "4/9/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Load required libraries

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(purrr)
library(readr)
library(tidyr)
library(ggplot2)
library(riingo)
options(pillar.sigfig = 6)
```

### Load data

```{r}
end_date <- as.Date("2018-03-30")
start_date <- end_date - years(8) + days(21) # Adjusted to be exactly 2001 points

# prices_raw <- tidyquant::tq_get(c("AAPL", "SPY"), from = start_date, to = end_date)
# prices_raw <- prices_raw %>% rename(ticker = symbol)
# write_rds(prices_raw, "prices_raw.rds")
prices_raw <- read_rds("prices_raw.rds")
```

### Relative shifts

The close prices that we get from Yahoo are nicely already adjusted for splits,
but not for dividends, but that is okay.

```{r, fig.width=10, fig.height=6}
prices_raw %>%
  select(ticker, date, close) %>%
  ggplot(aes(x = date, y = close, color = ticker)) +
  geom_line() +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2")
```

Here we calculate relative shifts

```{r}
prices_shifts <- prices_raw %>%
  group_by(ticker) %>%
  mutate(losses = (close / lag(close) - 1) * last(close)) %>% 
  select(ticker, date, losses) %>%
  na.omit() %>%
  spread(ticker, losses)

prices_shifts
```

### VaR calculations

```{r}
shares_aapl <- 200
shares_spy  <- 100

i <- seq_len(40)
alpha <- 0.0005 * i

cutoff <- floor(2000 * alpha)

calc_VaR <- function(losses, cutoff) {
  sort(losses)[cutoff]
}

historical_VaR <- tibble(alpha, cutoff) %>%
  mutate(
    VaR_AAPL = calc_VaR(shares_aapl * prices_shifts$AAPL, cutoff),
    VaR_SPY  = calc_VaR(shares_spy  * prices_shifts$SPY,  cutoff),
    VaR_port = calc_VaR(shares_spy  * prices_shifts$SPY + shares_aapl * prices_shifts$AAPL,  cutoff)
  )

historical_VaR
```

### Check for violations of coherence

I use the convention that a negative VaR is a bad thing. This is the same as 
saying that I use "negative losses" all throughout. Because of this, we check
the opposite condition of `VaR_port > VaR_AAPL + VaR_SPY`. If this condition
is true, we do NOT have a violation of coherence.

```{r}
violations <- historical_VaR %>%
  mutate(
    violation = ! (VaR_port > VaR_AAPL + VaR_SPY)
  )

violations
```

It does not look like any alpha level results in a violation

```{r}
# Does any alpha level result in a violation?
any(violations$violation == TRUE)
```

We can also check this with a visualization

```{r, fig.width=10, fig.height=8}
violations %>% 
  mutate(VaR_AAPL_plus_SPY = VaR_AAPL + VaR_SPY,
         alpha = factor(alpha)) %>%
  select(alpha, VaR_AAPL_plus_SPY, VaR_port) %>%
  gather("VaR_type", "VaR", contains("VaR")) %>%
  ggplot(aes(x = alpha, y = VaR, fill = VaR_type)) +
  geom_col(position = position_dodge(), width = .5) +
  coord_flip() +
  scale_y_continuous(expand = c(.01, 0)) +
  scale_x_discrete(position = "top") +
  scale_fill_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(hjust = .5), plot.subtitle = element_text(hjust = .5)) +
  ggtitle("Visually, we see the correct relationship holds. Portfolio VaR is less negative.", 
          subtitle = "VaR_port > VaR_AAPL + VaR_SPY")
```

