---
title: "HW 8"
author: "Davis Vaughan"
date: "3/28/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load data and libraries needed

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(purrr)
library(readr)
library(tidyr)
library(ggplot2)
library(knitr)
options(pillar.sigfig = 6)

portfolio <- c("AAPL", "SPY")

# Dates
end_date   <- as.Date("2018-01-05")
start_date <- end_date - years(2) + days(2) # Add two days to get the correct 504 prices

# library(tidyquant)
# portfolio_raw  <- tq_get(portfolio,  from = start_date, to = end_date, get = "stock.prices.google")
# write_rds(portfolio_raw, "portfolio_raw.rds")
portfolio_raw <- read_rds("portfolio_raw.rds")
```

# Question 2

### Absolute shifts

```{r}
portfolio_abs_shifts <- portfolio_raw %>%
  group_by(symbol) %>%
  mutate(delta_X = close - lag(close)) %>%
  select(symbol, date, delta_X) %>%
  na.omit()

portfolio_abs_shifts
```

### Custom multi lag function

You can really just ignore this.

Helpers

```{r}
# For each n, create a named list element holding the call
map_quos <- function(var, indices, default, lag_lead) {

  # For each index to lag/lead, create a quo call
  if(lag_lead == "lag") {
    lag <- dplyr::lag
    quo_list <- purrr::map(indices, ~rlang::quo(lag(!!var, !!.x, !!default)))
  } else if(lag_lead == "lead") {
    lead <- dplyr::lead
    quo_list <- purrr::map(indices, ~rlang::quo(lead(!!var, !!.x, !!default)))
  }

  # Create and set names
  fmt <- paste0("%s_", lag_lead, "_%02d")
  nms <- sprintf(fmt, as.character(var), indices)

  rlang::set_names(quo_list, nms)
}

extract_vars <- function(x, ...) {
  vars_char <- tidyselect::vars_select(colnames(x), !!! rlang::enquos(...))
  vars_syms <- rlang::syms(unname(vars_char))
  vars_syms
}

# A glue version of stop()
glue_stop <- function(..., .sep = "") {
  stop(glue::glue(..., .sep, .envir = parent.frame()), call. = FALSE)
}
```

Multi lag. This function lets you calculate multiple lags all at once.

```{r}
lags <- function(x, n = 1, ..., default = NA) {
  UseMethod("lags")
}

lags.default <- function(x, n = 1, ..., default = NA) {
  classes <- paste(class(x), collapse = ", ")
  glue_stop("No supported method of lags() for class: ", classes)
}

lags.data.frame <- function(x, n = 1, ..., default = NA){
  vars <- extract_vars(x, ...)

  # For each var to lag, create the list of calls
  calls_list <- purrr::map(vars, ~map_quos(.x, n, default, "lag"))

  # The above results in a list of lists. Flatten to get 1 list of calls
  calls <- rlang::flatten(calls_list)

  dplyr::mutate(x, !!! calls)
}

lags.tbl <- lags.data.frame
```

Example:

```{r}
lags(portfolio_abs_shifts, 1:2, delta_X)
```


### Correlogram setup

```{r}
correlate_lagged <- function(x, n = 1L) {
  x %>%
    lags(n, delta_X) %>%
    select(symbol, contains("delta_X")) %>%
    nest() %>%
    mutate(cor = map(data, ~corrr::correlate(.x, use = "complete.obs") %>% corrr::focus(delta_X))) %>%
    unnest(cor) %>%
    separate(rowname, c("name", "lag"), sep = -2, convert = TRUE) %>%
    mutate(lag = as.factor(lag))
}
```

```{r}
all_lags <- correlate_lagged(portfolio_abs_shifts, 1:10)

all_lags %>%
  select(-name) %>%
  rename(correlation_to_lag_0 = delta_X) %>%
  kable()
```

### Absolute shifts

There does not look to be any autocorrelation here.

```{r, fig.height=10, fig.width=10}
n <- filter(portfolio_abs_shifts, symbol == "AAPL") %>% nrow()
ci <- qnorm( (1 + .95) / 2 ) / sqrt(n)

all_lags %>%
  ggplot(aes(x = lag, y = delta_X)) +
  geom_col(width = .3) +
  geom_hline(yintercept = ci, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = -ci, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 0, size = .2) +
  facet_wrap(~symbol, nrow = 2) +
  coord_cartesian(ylim = c(-.2, 1)) +
  ggtitle("Correlogram - Absolute shifts")
```

### Absolute value of absolute shifts

It looks like there is autocorrelation in the absolute value of the shifts.

```{r, fig.height=10, fig.width=10}
portfolio_abs_shifts %>%
  mutate(delta_X = abs(delta_X)) %>%
  correlate_lagged(1:10) %>%
  ggplot(aes(x = lag, y = delta_X)) +
  geom_col(width = .3) +
  geom_hline(yintercept = ci, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = -ci, linetype = "dashed", color = "blue") +
  geom_hline(yintercept = 0, size = .2) +
  facet_wrap(~symbol, nrow = 2) +
  coord_cartesian(ylim = c(-.2, 1)) +
  ggtitle("Correlogram - Absolute value of absolute shifts")
```

### Ljung-Box test

We need a function to calculate the statistic and p-value.

```{r}
ljung_box <- function(x, h = 1) {
  
  stopifnot(h > 0)
  
  # Calculate the k-th box element
  box_k <- function(x, k, n) {
    autocor <- cor(x, lag(x, k), use = "pairwise.complete.obs")
    (autocor ^ 2) / (n - k)
  }
  
  n <- length(x)
  k_vals <- seq_len(h)
  Q_unadjusted <- sum(purrr::map_dbl(k_vals, ~box_k(x, .x, n)))
  Q <- n * (n + 2) * Q_unadjusted
  
  p_val <- 1 - pchisq(Q, df = h)
  
  tibble(Q = Q, p_val = p_val)
}
```

Fail to reject that the data are iid.

```{r}
portfolio_abs_shifts %>% 
  nest() %>%
  mutate(ljung = map(data, ~ljung_box(.$delta_X, 10))) %>%
  unnest(ljung) %>%
  select(symbol, Q, p_val) %>%
  kable()
```

Definitely reject that the absolute value of the shifts are iid.

```{r}
portfolio_abs_shifts %>% 
  mutate(delta_X_abs = abs(delta_X)) %>%
  nest() %>%
  mutate(ljung = map(data, ~ljung_box(.$delta_X_abs, 10))) %>%
  unnest(ljung) %>%
  select(symbol, Q, p_val) %>%
  kable()
```

### Summary

The shifts themselves don't seem to exhibit serial correlation, but the variance
of the shifts does. This is shown through the absolute value of the shifts showing
large amounts of serial correlation. Banks should likely be using _conditional
variance_ VaR models rather than purely unconditional models.

